---
title: "Bivariate Hypothesis Testing"
author: "Olivia Dobbs"
format: html
editor: visual
---

## Setup

```{r}
# Adding libraries for basic visualizations
library(ggplot2) 
library(gtsummary) 
library(tidyverse)
library(ggridges) 
library(janitor) 
library(rstatix)
library(gt)

# Adding loading/datasets for example suse
# Palmer penguins
library(palmerpenguins) # penguin measurements
penguins <- palmerpenguins::penguins
head(penguins)
```

### Two-Sample T-Test

**When:** continuous outcome variable, two independent groups in predictor, and independence between observations.

**Assumptions:** normal distribution of outcomes in each group, and variances in each group are similar.

**Null Hypothesis:** The population difference in mean bill length (mm) between the female and male groups of penguins is 0.

$H_0: \mu_{female} - \mu_{male} = 0$

**Alternative Hypothesis:** The population difference in mean bill length (mm) between the female and male groups of penguins is statistically different from 0.

$H_0: \mu_{female} - \mu_{male} \ne 0$

```{r}
# Cleaning the dataset for sake of example
clean_penguins = na.omit(penguins)

# Assuming independence of observations
# Observing normal distribution of groups
ggplot(data = clean_penguins,
       aes(x = bill_length_mm,
           y = sex,
           fill = sex)) +
  geom_density_ridges(alpha = 0.5, # changes opacity
                      show.legend = FALSE # removes the key to the right
  )
# Using the Central Limit Theorem as n for each group is > 164
# Checking variances
sumstats <- clean_penguins %>% 
  group_by(sex) %>% 
  get_summary_stats(type = "mean_sd") 
sumstats %>% gt()

# Performing the test
t.test(bill_length_mm ~ sex, data = penguins, var.equal = TRUE)
```

Results: p-value = 1.094e-10

Using alpha = 0.05, there is sufficient evidence that the population difference in mean bill length (mm) between male and female penguins is different from 0 (p \< 0.001). The mean bill length was 42.1 mm (SD = 4.9 mm) and 45.9 mm (SD = 5.4 mm) for the female and male groups, and the difference of 3.8 mm was statistically different.

### Wilcoxon Rank-Sum (Mann-Whitney)

### Paired T-Test

### Wilcoxon Signed-Rank

### One-Way ANOVA

### Kruskal-Wallis

### Pearson Correlation

### Spearman Correlation

### Chi-square Test of Independence

### Fisher's Exact Test

### McNemar's Test

### Exact Nemar
