---
title: "Bivariate Hypothesis Testing"
author: "Olivia Dobbs"
format: html
editor: visual
---

## Setup

```{r}
# Adding libraries for basic visualizations
library(ggplot2) 
library(gtsummary) 
library(tidyverse)
library(ggridges) 
library(janitor) 
library(rstatix)
library(gt)

# Adding loading/datasets for example suse
# Palmer penguins
library(palmerpenguins) # penguin measurements
penguins <- palmerpenguins::penguins
head(penguins)
```

### Two-Sample T-Test

**When:** continuous outcome variable, two independent groups in predictor, and independence between observations.

**Assumptions:** normal distribution of outcomes in each group, and variances in each group are similar.

**Null Hypothesis:** The population difference in mean bill length (mm) between the female and male groups of penguins is 0.

$H_0: \mu_{female} - \mu_{male} = 0$

**Alternative Hypothesis:** The population difference in mean bill length (mm) between the female and male groups of penguins is statistically different from 0.

$H_0: \mu_{female} - \mu_{male} \ne 0$

```{r}
# Cleaning the dataset for sake of example
clean_penguins = na.omit(penguins)

# Assuming independence of observations
# Observing normal distribution of groups
ggplot(data = clean_penguins,
       aes(x = bill_length_mm,
           y = sex,
           fill = sex)) +
  geom_density_ridges(alpha = 0.5, # changes opacity
                      show.legend = FALSE # removes the key to the right
  )
# Using the Central Limit Theorem as n for each group is > 164
# Checking variances
sumstats <- clean_penguins %>% 
  group_by(sex) %>% 
  get_summary_stats(type = "mean_sd") 
sumstats %>% gt()

# Performing the test
t.test(bill_length_mm ~ sex, data = penguins, var.equal = TRUE)
```

Results: p-value = 1.094e-10

Using alpha = 0.05, there is sufficient evidence that the population difference in mean bill length (mm) between male and female penguins is different from 0 (p \< 0.001). The mean bill length was 42.1 mm (SD = 4.9 mm) and 45.9 mm (SD = 5.4 mm) for the female and male groups, and the difference of 3.8 mm was statistically different.

### Wilcoxon Rank-Sum (Mann-Whitney)

```{r}
# Assuming that the distribution isn't normal enough for both groups for sake of example:

wilcox.test(bill_length_mm ~ sex, data = penguins)
```

### Paired T-Test

t.test(extra \~ group, data = sleep, paired = TRUE)

```{r}
# Adding a dataset that has data that's worth applying a paired t-test to:
# Same 10 IDs were tested with both drug types.
head(sleep)

# Observing data in a plot
boxplot(extra ~ group, data = sleep, 
        main = "Soporific Drug Effectiveness",
        xlab = "Drug Group", 
        ylab = "Extra Hours of Sleep",
        col = c("lightblue", "lightpink"))

# And performing the paired t-test
with(sleep, t.test(extra[group == 1], extra[group == 2], paired = TRUE))
# Note: "with()" is called an environment wrapper. It tells R to look inside 'sleep' for the variables that are mentioned inside of it. This option makes it so we don't have to type sleep$extra each time. 
# "extra[group == 1]" tells R "the 'extra' variable, where the condtion group == 1 is TRUE"
```

### Wilcoxon Signed-Rank

```{r}
# Graphing distribution of differences
sleep_diff_tidy <- sleep %>%
  mutate(group = paste0("drug_", group)) %>% # Renames 1 and 2 to drug_1 and drug_2
  pivot_wider(names_from = group, values_from = extra) %>%
  mutate(difference = drug_2 - drug_1)

ggplot(sleep_diff_tidy, aes(x = difference)) +
  geom_histogram(aes(y = after_stat(density)), bins = 5, 
                 fill = "skyblue", color = "white", alpha = 0.7) +
  geom_density(color = "darkblue", linewidth = 1) +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed", linewidth = 1) +
  labs(
    title = "Distribution of Sleep Differences",
    subtitle = "Values to the right of the red line mean Drug 2 outperformed Drug 1",
    x = "Difference in Hours (Drug 2 - Drug 1)",
    y = "Density"
  ) +
  theme_minimal()

# Assuming the differences between ours pairs are NOT NORMALLLY DISTRIBUTED
with(sleep, wilcox.test(extra[group == 1], extra[group == 2], paired = TRUE, exact = FALSE))
# Applying exact = FALSE because patient 5 showed 0 change between drugs and patients 3 and 4 had the exact same difference between their drugs. 
```

### One-Way ANOVA

### Kruskal-Wallis

### Pearson Correlation

### Spearman Correlation

### Chi-square Test of Independence

### Fisher's Exact Test

### McNemar's Test

### Exact Nemar
