---
title: "Bivariate Hypothesis Testing"
author: "Olivia Dobbs"
format: html
editor: visual
---

## Setup

```{r}
# Adding libraries for basic visualizations
library(ggplot2) 
library(gtsummary) 
library(tidyverse)
library(ggridges) 
library(janitor) 
library(rstatix)
library(gt)
library(exact2x2)

# Adding loading/datasets for example suse
# Palmer penguins
library(palmerpenguins) # penguin measurements
penguins <- palmerpenguins::penguins
head(penguins)
```

### Two-Sample T-Test

**When:** continuous outcome variable, two independent groups in predictor, and independence between observations.

**Assumptions:** normal distribution of outcomes in each group, and variances in each group are similar.

**Null Hypothesis:** The population difference in mean bill length (mm) between the female and male groups of penguins is 0.

$H_0: \mu_{female} - \mu_{male} = 0$

**Alternative Hypothesis:** The population difference in mean bill length (mm) between the female and male groups of penguins is statistically different from 0.

$H_0: \mu_{female} - \mu_{male} \ne 0$

```{r}
# Cleaning the dataset for sake of example
clean_penguins = na.omit(penguins)

# Assuming independence of observations
# Observing normal distribution of groups
ggplot(data = clean_penguins,
       aes(x = bill_length_mm,
           y = sex,
           fill = sex)) +
  geom_density_ridges(alpha = 0.5, # changes opacity
                      show.legend = FALSE # removes the key to the right
  )
# Using the Central Limit Theorem as n for each group is > 164
# Checking variances
sumstats <- clean_penguins %>% 
  group_by(sex) %>% 
  get_summary_stats(type = "mean_sd") 
sumstats %>% gt()

# Performing the test
t.test(bill_length_mm ~ sex, data = penguins, var.equal = TRUE)
```

Results: p-value = 1.094e-10

Using alpha = 0.05, there is sufficient evidence that the population difference in mean bill length (mm) between male and female penguins is different from 0 (p \< 0.001). The mean bill length was 42.1 mm (SD = 4.9 mm) and 45.9 mm (SD = 5.4 mm) for the female and male groups, and the difference of 3.8 mm was statistically different.

### Wilcoxon Rank-Sum (Mann-Whitney)

**When:** continuous outcome variable, two independent groups in predictor, and independence between observations.

**Assumptions:** distribution of outcomes in each group violates normality assumptions, and variances in each group are similar.

-   **Note:** If variances are not similar, the test can still be run, but it no longer showcases median difference. Significance where variance isn't similar means that there is **stochastic dominance**.

**Null Hypothesis:** The population difference in median bill length (mm) between the female and male groups of penguins is 0.

$H_0: M_{female} - M_{male} = 0$

**Alternative Hypothesis:** The population difference in median bill length (mm) between the female and male groups of penguins is statistically different from 0.

$H_0: M_{female} - M_{male} \ne 0$

```{r}
# Assuming that the distribution isn't normal enough for both groups for sake of example:

wilcox.test(bill_length_mm ~ sex, data = penguins)

# Getting median and IQR for results sentence
median(penguins$bill_depth_mm[penguins$sex == "male"], na.rm = TRUE)
IQR(penguins$bill_depth_mm[penguins$sex == "male"], na.rm = TRUE)

median(penguins$bill_depth_mm[penguins$sex == "female"], na.rm = TRUE)
IQR(penguins$bill_depth_mm[penguins$sex == "female"], na.rm = TRUE)
```

Results: p-value = 9.901e-11

Using alpha = 0.05, there is sufficient evidence that the population difference in median bill length (mm) between male and female penguins is different from 0 (p \< 0.001). The median bill length was 18.45 mm (IQR = 3.175 mm) and 17 mm (IQR = 3.3 mm) for the female and male groups, and the difference of 1.45 mm was statistically significant.

### Paired T-Test

```{r}
# Adding a dataset that has data that's worth applying a paired t-test to:
# Same 10 IDs were tested with both drug types.
head(sleep)

# Observing data in a plot
boxplot(extra ~ group, data = sleep, 
        main = "Soporific Drug Effectiveness",
        xlab = "Drug Group", 
        ylab = "Extra Hours of Sleep",
        col = c("lightblue", "lightpink"))

# And performing the paired t-test
with(sleep, t.test(extra[group == 1], extra[group == 2], paired = TRUE))
# Note: "with()" is called an environment wrapper. It tells R to look inside 'sleep' for the variables that are mentioned inside of it. This option makes it so we don't have to type sleep$extra each time. 
# "extra[group == 1]" tells R "the 'extra' variable, where the condtion group == 1 is TRUE"
```

### Wilcoxon Signed-Rank

```{r}
# Graphing distribution of differences
sleep_diff_tidy <- sleep %>%
  mutate(group = paste0("drug_", group)) %>% # Renames 1 and 2 to drug_1 and drug_2
  pivot_wider(names_from = group, values_from = extra) %>%
  mutate(difference = drug_2 - drug_1)

ggplot(sleep_diff_tidy, aes(x = difference)) +
  geom_histogram(aes(y = after_stat(density)), bins = 5, 
                 fill = "skyblue", color = "white", alpha = 0.7) +
  geom_density(color = "darkblue", linewidth = 1) +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed", linewidth = 1) +
  labs(
    title = "Distribution of Sleep Differences",
    subtitle = "Values to the right of the red line mean Drug 2 outperformed Drug 1",
    x = "Difference in Hours (Drug 2 - Drug 1)",
    y = "Density"
  ) +
  theme_minimal()

# Assuming the differences between ours pairs are NOT NORMALLLY DISTRIBUTED
with(sleep, wilcox.test(extra[group == 1], extra[group == 2], paired = TRUE, exact = FALSE))
# Applying exact = FALSE because patient 5 showed 0 change between drugs and patients 3 and 4 had the exact same difference between their drugs. 
```

### One-Way ANOVA

For performing a continuous outcome between 3+ groups.

```{r}
# Checking distribution and sample size
ggplot(data = penguins,
       aes(x = bill_length_mm,
           color = species)) + # Note that we put 'color' here instead of 'y'
  geom_density() 

table(penguins$species)

# Performing an anova, given n > 60 for each group and normal-appearing distributions
peng_anova <- aov(bill_depth_mm ~ species,
                  data = penguins)
summary(peng_anova)
```

**///TO DO: Add comparisons to determine which group(s) is/are significantly different from the others.**

### Kruskal-Wallis

For performing a continuous outcome between 3+ groups when anova assumptions are violated.

```{r}
# Assuming these species' bill depths violate normal distribution
# From the graph above and sample sizes, this isn't the case, but performing this test as an example. 
kruskal.test(bill_depth_mm ~ species,
             data = penguins)
```

### Pearson Correlation

For measuring the strength of a linear relationship between two quantitative & continuous variables.

```{r}
# Visualizing the relationship
ggplot(data = penguins,
       aes( x = bill_depth_mm,
            y = body_mass_g)) +
  geom_point() +
  geom_smooth(method = "lm")

# Testing pearson correlation
cor.test(~ bill_depth_mm + body_mass_g, 
         method = "pearson", 
         data = penguins)

# Note that this is a Simpson's Paradox. The species groups each have a positive correlation between bill depth and body mass, but together the correlation appears negative.
```

### Spearman Correlation

For measuring the strength of a linear relationship between two continuous variables. Unlike Pearson, however, Spearman is better for monotonic relationships, ordinal data, or skewed data distributions.

```{r}
data("USJudgeRatings")
#We'll be comparing demeanor vs integrity of judges in the US.

# Visualizing the relationship
ggplot(data = USJudgeRatings,
       aes( x = INTG,
            y = DMNR)) +
  geom_point() +
  geom_smooth(method = "lm")

# Testing pearson correlation
cor.test(USJudgeRatings$INTG, USJudgeRatings$DMNR, method = "spearman")
```

### Chi-square Test of Independence

For comparing the frequencies of two categorical variables

```{r}
data("HairEyeColor")
# This is a dataset that looks at the hair and eyecolor of a sample of statistics students
head(HairEyeColor)
# The data needs to be collapsed into a single table that isn't split by sex
hair_eye <- margin.table(HairEyeColor, margin = c(1,2))
head(hair_eye)

# Then we can run the chisq test to see if hair and eye color are independent.
chisq.test(hair_eye)
```

### Fisher's Exact Test

For comparing the frequencies of two categorical variables when expected cell counts are \<5

```{r}
# We can subset the hair-eye dataset to get smaller expected counts in our cells for an example.
hair_eye_small <- HairEyeColor[c("Black", "Blond"), c("Green", "Hazel"), "Female"]
head(hair_eye_small)

# Running a chisq.test() on a table that's too small will return a warning
chisq.test(hair_eye_small)

# In this case, we can then run a Fisher's Exact Test instead!
fisher.test(hair_eye_small)
```

### McNemar's Test

For paired data with a categorical outcome.

```{r}
# Making an example dataset for this option:
sentiment <- matrix(c(5254, 899, 348, 8881),
                    nrow = 2,
                    dimnames = list(
                      "Before Class" = c("Understanding", "Confused"),
                      "After Class" = c("Understanding", "Confused")
                    ))
sentiment

# Running the test
mcnemar.test(sentiment)
```

### Exact McNemar

For paired data with a categorical outcome, but when discordant pairs (understanding -\> confused and vice versa) add up to less than 25

```{r}
# Making an example dataset for this option:
sentiment_xs <- matrix(c(5, 8, 3, 8),
                    nrow = 2,
                    dimnames = list(
                      "Before Class" = c("Understanding", "Confused"),
                      "After Class" = c("Understanding", "Confused")
                    ))
sentiment_xs

mcnemar.exact(sentiment_xs)
```
