---
title: "Data Visualizations Example"
author: "Olivia Dobbs"
format: html
editor: visual
---

## Packages

```{r}
# Adding libraries for basic visualizations
library(ggplot2) # graphic representation for plots
library(gtsummary) # for generating presentation-ready tables
library(tidyverse) # inclusive package list for data science/manipulation/visualization
library(ggridges) # package to create ridgeline plots
library(janitor) # packing that helps with making things neat and neatly displaying data

# Adding loading/datasets for example suse
# Palmer penguins
library(palmerpenguins) # penguin measurements
penguins <- palmerpenguins::penguins
```

## Observing Datasets

Before plotting data, it helps to view summaries of the data you're working with

```{r}
# glimpse, from dplyr package in tidyverse
# shows first several data points for each variable
# shows the number of rows and columns in the dataset
glimpse(penguins)
```

```{r}
# summary gives summary information about variables included in the dataset
# includes minimum and maximum, and number of each in categories included
summary(penguins)
```

```{r}
# tbl_summary provides basic information on distributions of values in variables
tbl_summary(penguins)
```

```{r}
# use head to view the first 6 rows of a dataset
head(penguins)

# and tail shows the last few rows of a dataset
# you can specify how many rows are shown using a numerical parameter
tail(penguins, 10)
```

#### Frequency Tables

```{r}
# With a single variable
penguins %>% count(species)

# We can use tabyl for a prettier version
penguins %>% tabyl(species) %>%
  adorn_totals("row") %>% # adds a total at the bottom of the table 
  adorn_pct_formatting(digits=2) # allows the specification of significant figures for percent
```

#### Contingency Tables

```{r}
# With multiple variables at once
penguins %>% count(species, sex)

# Or a prettier version with tabyl
penguins %>% tabyl(species) %>%
  adorn_totals(c("row","col")) # totals calculated for each row and each column
```

#### Summarizing Averages

```{r}

# summary measurements like mean will return NA when NA values are present in a column.
# We can fix this by omitting the NA variables for now
clean_penguins = na.omit(penguins)

clean_penguins %>%
  summarize(
    mean = mean(bill_length_mm),
    sd = sd(bill_length_mm),
    median = median(bill_length_mm),
    min = min(bill_length_mm),
    max = max(bill_length_mm)
  )

# We can also group this by a categorical variable

clean_penguins %>%
  group_by(sex) %>%
  summarize(
    mean = mean(bill_length_mm),
    sd = sd(bill_length_mm),
    median = median(bill_length_mm),
    min = min(bill_length_mm),
    max = max(bill_length_mm)
  )
```

#### Making a Table 1

```{r}
# TO DO
clean_penguins %>%
  select( bill_length_mm, bill_depth_mm, flipper_length_mm, island,
          species, body_mass_g, sex, year) %>%
  tbl_summary(by = species,
              label = c(
                bill_length_mm ~ "Length of Bill (mm)",
                        bill_depth_mm ~ "Depth of Bill (mm)",
                        flipper_length_mm ~ "Length of Flippers (mm)",
                        island ~ "Island Measured From",
                        body_mass_g ~ "Body Mass (g)",
                        sex ~ "Sex",
                        year ~ "Year Measured"
                        ),
              statistic = list(all_continuous() ~ "{mean} ({sd})"),
              missing_text = "Unknown"
              )
```

# Plots

## GGPlot Plots

### Single Variables

#### Histograms

```{r}
# Basic historgram showing count
ggplot(data = penguins, # specifying what dataset to use
       aes(x = bill_length_mm)) + # assigning x variable in plot
       geom_histogram() # specifying what type of graph

# Basic histogram showing proportions in decimals
ggplot(data = penguins,
       aes(x = bill_length_mm)) + 
       geom_histogram(aes(y = stat(density)))

# Basic histogram showing proportions in percents
ggplot(data = penguins,
       aes(x = bill_length_mm)) + 
       geom_histogram(aes(y = stat(density))) +
  scale_y_continuous(labels = scales::percent_format())
```

#### Density Plot

```{r}
# Basic density plot
ggplot(data = penguins,
       aes(x = bill_length_mm)) +
       geom_density() 
```

#### Dot Plot

```{r}
ggplot(data = penguins,
       aes(x = bill_length_mm)) +
  geom_dotplot(binwidth = 0.5) + # binwidth sets the physical size of intervals used to group continuous data into bars on your plot
  labs(title = "Distribution of Bill Lengths", # labs lets you add text to your plot
       x = "Bill Lengths",
       y = "Count")
       
```

#### Boxplot

```{r}
ggplot(data = penguins,
       aes(x = bill_length_mm)) +
  geom_boxplot()

# for a vertical boxplot, you can also set bill length to 'y'
ggplot(data = penguins,
       aes(y = bill_length_mm)) +
  geom_boxplot()
```

#### Barplots

```{r}
ggplot(data = penguins,
       aes(x = species)) +
  geom_bar()
```

### Multiple Variables

#### Boxplots Grouped by A Variable

```{r}
ggplot(data = penguins,
       aes(x = bill_length_mm,
           y = species)) +
  geom_boxplot()
```

#### Jitter Graph Grouped by A Variable

```{r}
ggplot(data = penguins,
       aes(x = bill_length_mm,
           y = species)) +
  geom_jitter(
    aes(color = species)
  )
```

#### Combining Jitter Graph and Boxplot

```{r}
ggplot(data = penguins,
       aes(x = bill_length_mm,
           y = species)) +
  geom_boxplot() +
  geom_jitter(
    aes(color = species)
  )
```

#### Density Plots Grouped by A Variable

```{r}
ggplot(data = penguins,
       aes(x = bill_length_mm,
           color = species)) + # Note that we put 'color' here instead of 'y'
  geom_density() 
```

#### Ridgeline Plot

```{r}
ggplot(data = penguins,
       aes(x = bill_length_mm,
           y = species,
           fill = species)) +
  geom_density_ridges(alpha = 0.5, # changes opacity
                      show.legend = FALSE # removes the key to the right
  )
```

#### Scatterplots for Two Continuous Variables

```{r}
ggplot(data = penguins,
       aes( x = bill_length_mm,
            y = body_mass_g)) +
  geom_point()

# You can also add a categorical variable to color the points
ggplot(data = penguins,
       aes( x = bill_length_mm,
            y = body_mass_g,
            color = species)) +
  geom_point()

# You can also add a linear model line
ggplot(data = penguins,
       aes( x = bill_length_mm,
            y = body_mass_g,
            color = species)) +
  geom_point() +
  geom_smooth(method = "lm")
#-----------------------------------------------------------------------------------------------------------------------------------#
# Note that lm determines uncertainty bands by calculated a 95% CI for the predicted mean value of y at each specific value of x. 
# (It uses standard error of the fit - SE_y-hat)
# Instead of lm you can also use 
    # loess (smooth, flexible curve for non-linear patterns without assuming a pattern)
    # gam (general additive model that performs similar to loess, but better for n > 1000 datasets)
    # glm (assumes linear, but better for logistic regression or poisson regression)
    # rlm (a linear model that's less sensitive to outliers & needs the MASS library)
```

#### Bar Plots Between Two Categorical Variables

```{r}
ggplot(data = penguins,
       aes(x = species,
           fill = sex)) +
  geom_bar()

# We can also put them next to each other instead of stacked
ggplot(data = penguins,
       aes(x = species,
           fill = sex)) +
  geom_bar(position = "dodge")

```

## Base R Plots

#### Pie Chart

```{r}
# creating a dataframe of counts for penguin species
peng_species_count <- data.frame(
  group = c("Adelie", "Chinstrap", "Gentoo"),
  count = c(146,68,119)
)
pie(peng_species_count$count,
    labels = peng_species_count$group,
    main = "Species Share")
```

///// TO DO OTHER PLOTS TBD

////TO DO SAVING PLOTS
