---
title: "Data Management Example"
author: "Olivia Dobbs"
format: html
editor: visual
---

## Loading Dataset & Libraries

Loading all libraries necessary:

```{r}
library(dplyr)
library(tidyverse)
```

Loading csv and viewing

```{r}
exp_patient <- read_csv("../data/inpatient_patients_example.csv")

head(exp_patient)
```

[**NOTE:**]{.underline} **This data and the patients therein are randomly generated, and not based on any real people. Any patterns that emerge from this data set should not be used to make conclusions or draw patterns about any group or society.**

**THIS DATA SHOULD ONLY BE USED FOR THE PURPOSE OF LEARNING TO USE R.**

```{r}
table(exp_patient$icd9)
table(exp_patient$icd10)
```

## Step 1: Filtering by a CSV

Let's say we're interested in a sample that only includes our diagnoses categories of interest: cardiovascular and respiratory health issues.

```{r}
# importing a CSV that has all of the codes I'm interested in keeping for my analytic dataset.
icd_codes_cvd_resp <- read_csv("../data/cardiovascular_respiratory_icd_codes.csv")
# We usually want to have a chunk that imports everything in one place, but I'm keeping it here for illustrative purposes. 
# Note that if we only had a few ICD codes of interest, we could also manually create this object like this:
example_word_obj <- c("hello", "world", "here", "I", "am")


# creating an object that stores all the codes I'm interested in for this filter
cvd_resp <- (icd_codes_cvd_resp$code)

# filtering patient observations that have any numbers I'm interested in in either icd9 or icd10
exp_patient_clean <- exp_patient %>%
  filter(icd9 %in% cvd_resp | icd10 %in% cvd_resp)

# Examining exp_patient_clean to make sure it worked and observe changes from original dataset
nrow(exp_patient_clean) #289 is expected
head(exp_patient_clean)

```

## Step 2: Obtaining Length of Stay

Let's say we're interested in knowing how long our sample patients were at their respective hospitals before discharge:

```{r}
# First we need to calculate how long admission was for the patients
exp_patient_clean <- exp_patient_clean %>%
  mutate(
    admission_date = as.Date(admission_date),
    discharge_date = as.Date(discharge_date),
    los_days = as.numeric(discharge_date - admission_date)
  )

# And then visualizing to see our distribution of los
ggplot(data = exp_patient_clean,
       aes(x = los_days)) +
  geom_bar()
```

## Step 3: Exploring Category Collapsing

Let's say we're interested in decreasing the number of race/ethnicity categories and creating a single, race_eth variable.

```{r}
# First, we can explore the race and ethnicity variables
table(exp_patient_clean$race)
table(exp_patient_clean$ethnicity)
# And see how these categories overlap
table(exp_patient_clean$race, exp_patient_clean$ethnicity)

#creating a new variable
exp_patient_clean <- exp_patient_clean %>%
  mutate(
    race_eth = case_when(
      # Hispanic or Latino Category
      ethnicity == "Hispanic or Latino" &
        (race == "White" | race == "Some Other Race") ~
        "Hispanic or Latino",
      
      # Two Or More Races
      ethnicity == "Hispanic or Latino" &
        race != "White" & race != "Some Other Race" ~
        "Two or More Races",
      race == "Two or More Races" ~ "Two or More Races",
      
      # White
      (ethnicity != "Hispanic or Latino" | is.na(ethnicity)) &
        race == "White" ~
        "White",
      
      # Black
      (ethnicity != "Hispanic or Latino" | is.na(ethnicity)) &
        race == "Black or African American" ~
        "Black or African American",
      
      # AIAN & NHPI 
      (ethnicity != "Hispanic or Latino" | is.na(ethnicity)) &
        (race == "American Indian or Alaska Native" | 
           race == "Native Hawaiian or Other Pacific Islander") ~
        "AIAN or NHPI",
      
      # Asian
      (ethnicity != "Hispanic or Latino" | is.na(ethnicity)) &
        race == "Asian" ~ "Asian",
      
      # Other 
      (ethnicity != "Hispanic or Latino" | is.na(ethnicity)) &
        race == "Some Other Race" ~
        "Other"
    )
  )

# Looking at new variable
table(exp_patient_clean$race_eth)
```
