---
title: "Purposeful Model Selection Example"
author: "Olivia Dobbs"
format: html
editor: visual
---

## Packages and Data Import

```{r}
# libraries needed
library(tidyverse)
library(here)
library(readr)
library(gtsummary)
library(GGally)
library(gt)
library(broom)
library(gridExtra)

# dataset
exp <- read_csv(here("data", "population_bp_example.csv"))
glimpse(exp)
```

```{r}
head(exp)
```

[**Note:**]{.underline} The data for this example is entirely synthetic and generated for the purposes of this example. No results from this data should be considered indicative of any real patterns of health in any real population.

## Step 0: Defining a Research Question

##### How is systolic blood pressure (mmHg) associated with daily grams of sodium intake?

## Step 1: Exploring Simple Relationships

```{r}
exp %>% ggpairs(progress = FALSE)
```

#### SLR with main effect of interest and outcome

```{r}
mod_main_ef = exp %>%
  lm(formula = systolic_bp ~ sodium_intake_g_day)

tidy(mod_main_ef) %>% gt() %>%
  tab_options(table.font.size = 15) %>%
  fmt_number(decimals = 3)

# Borderline insignificant.
```

#### SLR with other variables of interest

```{r}
# This version is simpler and allows for more output control, but requires repetitive code.
# age
mod_age = exp %>%
  lm(formula = systolic_bp ~ age)

tidy(mod_age) %>% gt() %>%
  tab_options(table.font.size = 15) %>%
  fmt_number(decimals = 3)

# sex
mod_sex = exp %>%
  lm(formula = systolic_bp ~ sex)

tidy(mod_sex) %>% gt() %>%
  tab_options(table.font.size = 15) %>%
  fmt_number(decimals = 3)

# race_ethnicity
mod_eth = exp %>%
  lm(formula = systolic_bp ~ race_ethnicity)

tidy(mod_eth) %>% gt() %>%
  tab_options(table.font.size = 15) %>%
  fmt_number(decimals = 3)

# education
mod_edu = exp %>%
  lm(formula = systolic_bp ~ education)

tidy(mod_edu) %>% gt() %>%
  tab_options(table.font.size = 15) %>%
  fmt_number(decimals = 3)

# smoking_status
mod_smo = exp %>%
  lm(formula = systolic_bp ~ smoking_status)

tidy(mod_smo) %>% gt() %>%
  tab_options(table.font.size = 15) %>%
  fmt_number(decimals = 3)

# bmi
mod_bmi = exp %>%
  lm(formula = systolic_bp ~ bmi)

tidy(mod_bmi) %>% gt() %>%
  tab_options(table.font.size = 15) %>%
  fmt_number(decimals = 3)

# physical_activity_hrs_week
mod_phys = exp %>%
  lm(formula = systolic_bp ~ physical_activity_hrs_week)

tidy(mod_phys) %>% gt() %>%
  tab_options(table.font.size = 15) %>%
  fmt_number(decimals = 3)

# sleep_hrs_night
mod_zzz = exp %>%
  lm(formula = systolic_bp ~ sleep_hrs_night)

tidy(mod_zzz) %>% gt() %>%
  tab_options(table.font.size = 15) %>%
  fmt_number(decimals = 3)
```

```{r}
# Alternative, quick method with more efficient code

exp_int = exp %>%
  lm(formula = systolic_bp ~ 1)

add1(exp_int,
     scope = ~ age + sex + race_ethnicity + education + smoking_status + bmi + physical_activity_hrs_week + sodium_intake_g_day + sleep_hrs_night,
     test = "F")
```

#### Observing ANOVAs

```{r}
# This is an additional step we need to make in light of how we created SLR models in the previous chunks of this section. 
anova(mod_main_ef)
anova(mod_age)
anova(mod_bmi)
anova(mod_edu)
anova(mod_eth)
anova(mod_phys)
anova(mod_sex)
anova(mod_smo)
anova(mod_zzz)
```

## Step 2: Preliminary Variable Selection

Using values from the **add1** function above, I'm beginning model selection using all variables where the resulting p-value is less than 0.25.

As a result, this initial model will **not** include:

-   Weekly physical activity hours (0.410)

-   Hours of sleep each night (p = 0.346)

The remaining variables will be included.

```{r}
init_mod = exp %>%
  lm(formula = systolic_bp ~ sodium_intake_g_day + age + sex + race_ethnicity + education + smoking_status + bmi)


tbl_regression(
  init_mod) %>%
  as_gt() %>%
  tab_options(table.font.size = 15)
```

## Step 3: Assess change in coefficient

I am opting to further remove variables in cases where:

The p-value for the F-test of it's coefficients are \> 0.05

and the change in coefficient of our main variable of interest is \<10% with it's removal

```{r}
drop1(init_mod, test="F")

# Resulting variables worth considering a drop: education (p ~ 0.49)
# Displaying an ANOVA table to test
full_mod = init_mod
red_mod = lm(formula = systolic_bp ~ sodium_intake_g_day + age + sex + race_ethnicity + smoking_status + bmi,
             data = exp)

anova(full_mod, red_mod) %>% tidy() %>%
  gt() %>% tab_options(table.font.size = 15) %>% fmt_number(decimals = 3)

# Change in coefficient calculation:
# Beta-1 full = 1.1
tbl_regression(
  red_mod) %>%
  as_gt() %>%
  tab_options(table.font.size = 15)
# Beta-1 reduced = 1.1
```

With a small percent change in the main effect coefficient and a non-significant ANOVA model comparison, I opted to remove education from the final model. The extra variable does not significantly improve the fit of the model, and has a very small impact on the main effect.

Thus, I'll be proceeding with the reduced model.

## Step 4: Assess scale for continuous variables

Main Effects Model

```{r}
# Continuous variables are: sodium, age, & bmi

sodium = ggplot(data = exp, aes(y = systolic_bp, x = sodium_intake_g_day)) + 
  geom_point() +
  geom_smooth(se=F) + labs(x = "NA Intake (g/day)", y = "Systolic BP (mmHg)")

age = ggplot(data = exp, aes(y = systolic_bp, x = age)) + 
  geom_point() +
  geom_smooth(se=F) + labs(x = "Age", y = "Systolic BP (mmHg)")

bmi = ggplot(data = exp, aes(y = systolic_bp, x = bmi)) + 
  geom_point() +
  geom_smooth(se=F) + labs(x = "BMI", y = "Systolic BP (mmHg)")

grid.arrange(sodium, age, bmi, nrow=1)
```

For now, I'll proceed with keeping these 3 variables continuous and without transformation/splines as, though they're not *perfectly* linear, they are linear enough from a quick visual check. Residuals in later steps may help clarify whether some transformation is necessary.

## Step 5: Check for interactions

Variables with clinical plausibility to interact with daily sodium intake.

**Age:** salt sensitivity increases with age

**BMI:** BMI is associated with sodium intake

**Sex:** Hormone prevalence may impact salt's effect on vascular health

**Race:** Cultural differences in dietary sources may impact overall intake, and there are genetic differences in salt sensitivity between some racial groups.

**Smoking Status:** Smoking alters tastes preferences & increases salt cravings

```{r}
main_eff_model = red_mod # Relabeling for easier reference
add1(main_eff_model, scope = ~sodium_intake_g_day * . , test = "F")
# None of the interaction additions are significant, keeping none of them in the final model.
```

## Step 6: Assess model fit

Gathering Model Fit Statistics

NOTE: MLR Diagnostics will be performed in

```{r}
# Adjusted R-squared, AIC, and BIC
glance(main_eff_model) %>% mutate(Model = "Final model") %>%
  select(Model, adj.r.squared, AIC, BIC) %>%
  gt() %>% 
  tab_options(table.font.size = 15) %>%
  fmt_number(decimals = 3)
```

TO DO: Write a summary here that showcases:

-   "Final" model equation

-   "Final" conclusions to be drawn.

-   AND GENERALLY INTERPET RESULTS

-   AND CHECK EVERYTHING
