---
title: "Purposeful Model Selection Example"
author: "Olivia Dobbs"
format: html
editor: visual
---

## Packages and Data Import

```{r}
# libraries needed
library(tidyverse)
library(here)
library(readr)
library(gtsummary)
library(GGally)

# dataset
exp <- read_csv(here("data", "population_bp_example.csv"))
glimpse(exp)
```

```{r}
head(exp)-
```

[**Note:**]{.underline} The data for this example is entirely synthetic and generated for the purposes of this example. No results from this data should be considered indicative of any real patterns of health in any real population.

## Step 0: Defining a Research Question

##### How is systolic blood pressure (mmHg) associated with daily grams of sodium intake?

## Step 1: Exploring Simple Relationships

```{r}
exp %>% ggpairs(progress = FALSE)
```

#### SLR with main effect of interest and outcome

```{r}
mod_main_ef = exp %>%
  lm(formula = systolic_bp ~ sodium_intake_g_day)

tidy(mod_main_ef) %>% gt() %>%
  tab_options(table.font.size = 15) %>%
  fmt_number(decimals = 3)
```

#### SLR with other variables of interest

```{r}
# This version is simpler and allows for more output control, but requires repetitive code.
# age
mod_age = exp %>%
  lm(formula = systolic_bp ~ age)

tidy(mod_age) %>% gt() %>%
  tab_options(table.font.size = 15) %>%
  fmt_number(decimals = 3)

# sex
mod_sex = exp %>%
  lm(formula = systolic_bp ~ sex)

tidy(mod_sex) %>% gt() %>%
  tab_options(table.font.size = 15) %>%
  fmt_number(decimals = 3)

# race_ethnicity
mod_eth = exp %>%
  lm(formula = systolic_bp ~ race_ethnicity)

tidy(mod_eth) %>% gt() %>%
  tab_options(table.font.size = 15) %>%
  fmt_number(decimals = 3)

# education
mod_edu = exp %>%
  lm(formula = systolic_bp ~ education)

tidy(mod_edu) %>% gt() %>%
  tab_options(table.font.size = 15) %>%
  fmt_number(decimals = 3)

# smoking_status
mod_smo = exp %>%
  lm(formula = systolic_bp ~ smoking_status)

tidy(mod_smo) %>% gt() %>%
  tab_options(table.font.size = 15) %>%
  fmt_number(decimals = 3)

# bmi
mod_bmi = exp %>%
  lm(formula = systolic_bp ~ bmi)

tidy(mod_bmi) %>% gt() %>%
  tab_options(table.font.size = 15) %>%
  fmt_number(decimals = 3)

# physical_activity_hrs_week
mod_phys = exp %>%
  lm(formula = systolic_bp ~ physical_activity_hrs_week)

tidy(mod_phys) %>% gt() %>%
  tab_options(table.font.size = 15) %>%
  fmt_number(decimals = 3)

# sleep_hrs_night
mod_zzz = exp %>%
  lm(formula = systolic_bp ~ sleep_hrs_night)

tidy(mod_zzz) %>% gt() %>%
  tab_options(table.font.size = 15) %>%
  fmt_number(decimals = 3)
```

```{r}
# Alternative, quick method with more efficient code

exp_int = exp %>%
  lm(formula = systolic_bp ~ 1)

add1(exp_int,
     scope = ~ age + sex + race_ethnicity + education + smoking_status + bmi + physical_activity_hrs_week + sodium_intake_g_day + sleep_hrs_night,
     test = "F")
```

#### Observing ANOVAs

```{r}
anova(mod_main_ef)
anova(mod_age)
anova(mod_bmi)
anova(mod_edu)
anova(mod_eth)
anova(mod_phys)
anova(mod_sex)
anova(mod_smo)
anova(mod_zzz)
```

## Step 2: Preliminary Variable Selection

Using values from the add1 function above, I'm beginning model selection using all variables where the resulting p-value is less than 0.25.

As a result, this initial model will **not** include:

-   Weekly physical activity hours (0.410)

-   Hours of sleep each night (p = 0.346)

The remaining variables will be included.

```{r}
init_mod = exp %>%
  lm(formula = systolic_bp ~ sodium_intake_g_day + age + sex + race_ethnicity + education + smoking_status + bmi)


tbl_regression(
  init_mod) %>%
  as_gt() %>%
  tab_options(table.font.size = 15)
```

```{r}
drop1(init_mod, test="F")
```

## Step 3: Assess change in coefficient

```{r}
# Displaying an ANOVA table
```

## Step 4: Assess scale for continuous variables

Main Effects Model

## Step 5: Check for interactions

## Step 6: Assess model fit
