---
title: "Random Sampling Example"
author: "Olivia Dobbs"
format: html
editor: visual
---

## Library Setup

```{r}
library(survey)
library(sampling)
library(dplyr)
library(ggplot2)
library(here)
library(readr)
```

```{r}
# Adding in our example dataset
exp <- read_csv(here("data", "population_bp_example.csv"))
glimpse(exp)
```

## Example Sampling

```{r}
# Setting a seed for reproducibility. 
# Without this, the random selection will change each time it's run. 
set.seed(2026)
```

### DPLYR Version

```{r}
# If we want to sample a specified amount of rows
sampled_data_n <- slice_sample(exp, n = 10)

# If we want to sample a specific proportion of our data
sampled_data_prop <- slice_sample(exp, prop = 0.20)

# And then we can look at these two different samples:
glimpse(sampled_data_n)
glimpse(sampled_data_prop)
```

### Base R Version

```{r}
# Selecting a sample of 6 indices from my dataset, exp
# Sample 6 numbers anywhere from 1 to the length of dataset exp
random_indices <- sample(1:nrow(exp), 6) 


# Extracting the rows from the dataset, exp, that match the random indices
sampled_data <- exp[random_indices, ]
head(sampled_data)
```

## Example Calculation 1: SRS *Without* Replacement

When sampling from a population of a known, finite size, we can estimate information about that population by considering an outcome in our sample and applying a correction.

```{r}
## Information "given" by an experiment ##
# Population size from which we're sampling
N1 <- 55234

# Sample we had access to
n1 <- 500

# Number within sample who had our trait of interest
flag <- 44

## Calculations to use a finite-population correction for an SRS ##
# Proportion of those with flag in sample
prop <- flag/n1

# Determining weight
w <- 1/prop

# Computing the sample variance estimate for our binary flag variable
sSquare1 <- (n1/(n1-1))*(prop*(1-prop))

# Computing the standard error with the finite population correction for sampling without replacement
SEwor <- sqrt((1-(n1/N1))*(sSquare1/n1))

# Constructing the 95% CI
lb1 <- prop - qnorm(.975)*SEwor
ub1 <- prop + qnorm(.975)*SEwor

# And outputting the confidence interval...
c(lb1,ub1)
# the standard error...
SEwor
# and the proportion of sample with the flag
prop
```

From this calculation we could conclude...

-   "In a simple random sample without replacement, the estimated proportion of \_\_\_\_\_ who \_\_\_\_\_\_ was 0.088 (8.8%) with a standard error of 0.01262 (95% CI: 0.06326, 0.1127)"

## Example Calculation 2: SRS *With* Replacement

When a selected observation can be picked again following its sampling, we don't need to apply the finite-population correction.

```{r}
# Calculating a new standard error without the finite population correction
SEwr <- sqrt(sSquare1 / n1)

# Constructing our new 95% Confidence Interval
lb1 <- prop - qnorm(.975)*SEwr
ub1 <- prop + qnorm(.975)*SEwr

# Outputting our values
c(lb1,ub1)
SEwr
```

From this calculation we could conclude...

-   "In a simple random sample with replacement, the estimated proportion of \_\_\_\_\_ who \_\_\_\_\_\_ was 0.088 (8.8%) with a standard error of 0.01268 (95% CI: 0.06314, 0.1129)"
